---
title: "cleanup"
author: "Natalie Coy & Danny Foster"
date: "7/13/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}

#### install packages ##########################################################
library(tidyverse)
library(here)
library(readxl)
library(ggplot2)


#### load data #################################################################
#need to update filepath when there is new data imported 
filepath = 
  here("02-data", "00-source", "preburn", "CARB_datasheets_final.xls")

plot_metadata <- 
  read_excel(filepath, 
             sheet = "plot_metadata", 
             col_types = c("text", "text", "text", "numeric", "numeric", 
                           "numeric", "numeric", "numeric", "text", "numeric", 
                           "date", "numeric", "numeric", "numeric", "date"))

cwd <- 
  read_excel(filepath, 
             sheet="cwd", 
             col_types = c("text", "text", "text", "numeric", "numeric", 
                           "numeric", "numeric", "numeric", "text", "text"))

fwd_tallies <- 
  read_excel(filepath, 
             sheet="fwd_tallies", 
             col_types = c("text", "text", "text", "numeric", "numeric", 
                           "numeric", "numeric", "numeric", "numeric", "numeric"))
# NAs investigated during cleaning


fwd_diams <- 
  read_excel(filepath, 
             sheet="fwd_diams", 
             col_types = c("text", "text", "text", "numeric"))

litter_duff <- 
  read_excel(filepath, 
             sheet="litter_duff", 
             col_types = c("text", "text", "text", "numeric", "numeric", 
                           "numeric"))

shrubs <- 
  read_excel(filepath, 
             sheet="shrubs", 
             col_types = c("text", "text", "text", "numeric", "numeric", 
                           "text", "numeric", "numeric", "numeric", "text"))

trees <- 
  read_excel(filepath, 
             sheet="trees", col_types = c("text", "text", "numeric", 
                                          "numeric", "text", "numeric", 
                                          "numeric", "numeric", "numeric", 
                                          "text", "text", "text"))

pres_abs_2 <- 
  read_excel(filepath, 
             sheet = "pres_abs_2", 
             col_types = c("text", "text", "text", "numeric", "numeric", 
                           "numeric", "numeric", "numeric", "numeric"))

pres_abs <- 
  read_excel(filepath, 
             sheet = "pres_abs", 
             col_types = c("text", "text", "text", "numeric", "numeric", 
                           "numeric", "numeric", "numeric", "numeric"))


```




#Data Cleaning 

## Plot metadata

```{r}
head(plot_metadata)

plot_metadata = 
  plot_metadata %>%
  # construct the observation ids PLOT and TRANSECT, format the date nicely
  mutate(PLOT = paste0(COMP, "-", 
                       stringr::str_pad(PLOT, width = 3, side = "left", pad = 0))) %>%
  mutate(TRANSECT = paste0(PLOT, "-", str_pad(AZ, width = 3, side = "left", pad = 0)), 
         AZ = as.numeric(AZ),
         Date = as.Date(Date, format = '%Y-%m-%d')) %>%
  
  # keep only the most relevant columns
  select(COMP, PLOT, TRANSECT, Date, AZ, SLP_pcent, HerbCover, comment) %>%
  
  # add column for out-of-bounds distance (transect extended beyond compartment)
  mutate(max_dist_m = 30)


# manually update a few out-of-bounds distances based on reading the sheet
plot_metadata[plot_metadata$TRANSECT=='A-023-000','max_dist_m'] = 18.1
plot_metadata[plot_metadata$TRANSECT=='A-024-090','max_dist_m'] = 17.0
plot_metadata[plot_metadata$TRANSECT=='A-016-270','max_dist_m'] = 20.6
plot_metadata[plot_metadata$TRANSECT=='C-022-270','max_dist_m'] = 17.0

```

## Litter and Duff 

  - Litter and Duff depths should always be greater or equal to zero 
  - Only certain valid locations 
  - NA values okay in the Litter_cm and Duff_cm columns 
  - no NA in COMP, PLOT, AZ, or location 
  - COMP will be either A, B, or C 
  - set number of rows 
  - specific set of COMP/PLOT combinations that exist 
  - For each plot there should be 4 different AZ values, 0, 90, 270, 360
  - For each Comp/Plot/AZ value there should be a specific set of location values 
  - Check for high outliers on Litter and duff depth 
  

```{r}
head(litter_duff)

#Litter and Duff depths should be >= 0, this also checks for high outliers on litter and duff  
summary(litter_duff)

litter_duff %>% 
  filter(LITTER_cm<0 | DUFF_cm<0) 

#Only a specific set of locations 

valid_locations <- 
  c(2, 2.1, 2.2, 2.3, 2.4, 2.5, 3, 3.5, 4, 4.5, 5.5, 7, 7.5, 9.5, 19.5, 29.5)

litter_duff %>%
  filter(!is.element(LOCATION, valid_locations))

# NAs in litter and duff ok - locatinos with rocks/stumps/logs etc
litter_duff %>%
  filter(is.na(LITTER_cm) | is.na(DUFF_cm))

litter_duff %>%
  filter(is.na(COMP) | is.na(PLOT) | is.na(LOCATION))


#COMP will be either A, B, or C 

valid_COMP <- 
  c("A", "B", "C")

litter_duff %>%
  filter(!is.element(COMP, valid_COMP))

#set number of rows, there are 3 COMP with a total of 19 different plots, 
# with four transects per plot, with 16 locations on each transect 
#table failed the test, need to come back once the data is complete 
length(valid_locations)

valid_row_number <- 
  19 * 4 * 16 

valid_row_number == 
  nrow(litter_duff)

# there's 1 extra litter/duff sample, at 6.5m on A-21-0. I'm going to leave it in.
valid_row_number
nrow(litter_duff)

#specific COMP/PLOT/AZ/LOCATION values; creating ID versions which are 
# explicitly nested charater strings (TRANSECT, LOCATION), 
# and geometry versions (displacement_m, AZ) of the transect azimuth and 
# distance-along-transect for each observation. ID strings are to organize 
# observations, fit aspatial random effects, etc. Geometry versions will be 
# used to geolocate samples.
litter_duff = litter_duff %>%
  mutate(PLOT = paste0(COMP, "-", 
                       stringr::str_pad(PLOT, width = 3, side = "left", pad = 0))) %>%
  mutate(TRANSECT = paste0(PLOT, "-", str_pad(AZ, width = 3, side = "left", pad = 0)), 
         AZ = as.numeric(AZ), 
         Displacement_m = LOCATION, 
         LOCATION = paste0(TRANSECT, "-", LOCATION)) %>%
  select(COMP, PLOT, TRANSECT, LOCATION, AZ, Displacement_m, LITTER_cm, DUFF_cm)

#confirm all of our LOCATION - All of the correct combinations of COMP, PLOT, TRANSECT, LOCATION
valid_plots = c("A-001", "A-003", "A-005", "A-021", "A-023", "A-025", "A-020", "A-024", 
                "A-028", "B-008", "B-013", "B-010", "B-016", "B-026", "C-010", 
                "C-011", "C-025", "C-026", "C-022")

valid_locations <- 
  expand.grid(valid_plots, 
              c("000", "090", "180", "270"), 
              c(2, 2.1, 2.2, 2.3, 2.4, 2.5, 3, 3.5, 4, 4.5, 5.5, 7, 7.5, 9.5, 
                19.5, 29.5)) %>%
  select(PLOT = Var1, TRANSECT = Var2, Displacement_m = Var3) %>%
  mutate(LOCATION = paste0(PLOT, "-", TRANSECT, "-", Displacement_m)) 

#Selecting the location column from the table to use 
#view(valid_locations)
valid_locations = valid_locations$LOCATION

#Checking for extra locations; again just the extra sample at 6.5m
litter_duff %>%
  filter(!is.element(LOCATION, valid_locations))

#Checking for missing locations 
valid_locations[!is.element(valid_locations,litter_duff$LOCATION)]

```

## FWD Tallies

  - a1h, a10h, a100h, b1h, b10h, b100h columns should be >= 0 
  - Only certain valid locations 
  - NA values okay in the a1h, a10h, a100h, b1h, b10h, b100h columns 
  - no NA in COMP, PLOT, AZ, or location 
  - set number of rows 
  - specific set of COMP/PLOT combinations that exist 
  - For each plot there should be 4 different AZ values, 0, 90, 270, 360
  - For each Comp/Plot/AZ value there should be a specific set of location values 
  - Check for high outliers on a1h, a10h, a100h, b1h, b10h, b100h columns 

```{r}
head(fwd_tallies)
#rename 8 to COMP, was entered in incorrectly 
fwd_tallies = 
  fwd_tallies %>%
  select(COMP=`8`, PLOT, AZ, LOCATION, a1h, a10h, a100h, b1h, b10h, b100h)

#fwd_tallies should be >=0, this also checks for high outliers 
summary(fwd_tallies)

fwd_tallies %>%
  filter(a1h<0 | a10h<0 | a100h<0 | b1h<0| b10h<0 | b100h<0) 

#Only a specific set of locations 
valid_locations2 <-  
  c(2.5, 3.5, 4.5, 5.5, 7.5, 9.5, 19.5, 29.5)

fwd_tallies %>%
  filter(!is.element(LOCATION, valid_locations2))

#Checking for NAs; 7 NAs have been cleared; 2 missing 100h tallies, which are 
# PROBABLY 0s but can't be sure
#A- 023 - 000 : transect hit Comp boundary at 18.1m 
#A- 024 - 090 : transect hit Comp boundary at 17m
#A- 024 - 180 : data has not been entered into the spreadsheet 
#A- 024 - 270 : leave NA if there was a stump or rock on transect 
#B- 016 - 270 : transect hit comp boundary at 20.6m 
#C- 022 - 270 - 9.5 : transect got an NA due to a stump or rock 
#C- 022 - 270 - 19.5 and 29.5 : transect hit comp boundary at 17.0m 
fwd_tallies %>%
  filter(is.na(a1h) | is.na(a10h) | is.na(a100h) | is.na(b1h) | is.na(b10h) | is.na(b100h))



fwd_tallies %>%
 filter(is.na(COMP) | is.na(PLOT) | is.na(LOCATION) | is.na(AZ))

#COMP will be either A B or C 
valid_COMP <- 
  c("A", "B", "C")

fwd_tallies %>%
  filter(!is.element(COMP, valid_COMP))

#set number of rows, there are 3 COMP with a total of 18 different plots, with four transects per plot, with 8 locations on each transect 

valid_row_number <- 
  19 * 4 * 8 

valid_row_number == 
  nrow(fwd_tallies)

valid_row_number
nrow(fwd_tallies)

##specific COMP/PLOT/AZ/LOCATION values 
fwd_tallies = 
  fwd_tallies %>%
  mutate(PLOT = paste0(COMP, "-", 
                       stringr::str_pad(PLOT, width = 3, side = "left", pad = 0))) %>%
  mutate(TRANSECT = paste0(PLOT, "-", str_pad(AZ, width = 3, side = "left", pad = 0)), 
         AZ = as.numeric(AZ), 
         Displacement_m = LOCATION, 
         LOCATION = paste0(TRANSECT, "-", LOCATION)) %>%
  select(COMP, PLOT, TRANSECT, LOCATION, AZ, Displacement_m, a1h, a10h, a100h, b1h, b10h, b100h)

#confirm all of our LOCATION - All of the correct combinations of COMP, PLOT, AZ, Displacement_m
valid_locations2 <- 
  expand.grid(c("A-001", "A-003", "A-005", "A-023", "A-025", "A-020", "A-024", 
                "A-028", "B-008", "B-013", "B-010", "B-016", "B-026", "C-010", 
                "C-011", "C-025", "C-026", "C-022"), 
              c("000", "090", "180", "270"), 
              c( 2.5, 3.5, 4.5, 5.5, 7.5, 9.5, 19.5, 29.5)) %>%
  select(PLOT = Var1, TRANSECT = Var2, Displacement_m = Var3) %>%
  mutate(LOCATION = paste0(PLOT, "-", TRANSECT, "-", Displacement_m)) 

valid_locations2 = valid_locations2$LOCATION

#Checking for extra locations 
fwd_tallies %>%
  filter(!is.element(LOCATION, valid_locations2))

#Checking for missing locations 
valid_locations2[!is.element(valid_locations2,fwd_tallies$LOCATION)]

# there's one sample with a crazy high value for 100-hours:
fwd_tallies %>%
  filter(a100h>30)

# looks like a data entry typo, where they are missing one of the subsample a
# observations, and the 118 should be for b1h. Not sure what the missing 
# subsample is, just set this whole row to NA
fwd_tallies = 
  fwd_tallies %>%
  mutate(a1h = ifelse(a100h>30,NA,a1h),
         a10h = ifelse(a100h>30,NA,a10h),
         a100h = ifelse(a100h>30,NA,a100h),
         b1h = ifelse(a100h>30,NA,b1h),
         b10h = ifelse(a100h>30,NA,b10h),
         b100h = ifelse(a100h>30,NA,b100h))

# there's also some high values for 10h fuels, but these look legit. This is 
# the woodrat next I think.
fwd_tallies %>% 
  filter(a10h >30|b10h>30)
```

## FWD diams
  - DIAM_mm should be >= 0 
  - Should be no NA values in any columns 
  - COMP will be either A, B, or C 
  - specific set of COMP/PLOT combinations that exist 
  - For each plot there should be 4 different AZ values, 0, 90, 270, 360
  - Check for high outliers on DIAM_mm 
  
```{r}
head(fwd_diams)

#DIAM_mm should be >=0, this also checks for high outliers on fwd_diams 
#If diam is greater than 76.3mm is considered a 1000h fuel and would not be 
# included in this table, maximum of 74mm is okay in this case 
# identified one NA
#the row has a blank value, with no comment in metadata likely to be a recording error 
summary(fwd_diams)
hist(fwd_diams$DIAM_mm)

fwd_diams %>% 
  filter(DIAM_mm<0)

fwd_diams %>% 
  filter(is.na(DIAM_mm))

#decided to drop the NA value 
fwd_diams = 
  fwd_diams %>%
  filter(!is.na(DIAM_mm))

#COMP will either be A B or C 
valid_COMP <- 
  c("A", "B", "C")

fwd_diams %>%
  filter(!is.element(COMP, valid_COMP))

#Specific COMP/PLOTS/AZ
valid_transects <- 
  expand.grid(c("A-001", "A-003", "A-005","A-021", "A-023", "A-025", "A-020", "A-024", 
                "A-028", "B-008", "B-013", "B-010", "B-016", "B-026", "C-010", 
                "C-011", "C-025", "C-026", "C-022"), 
              c("000", "090", "180", "270")) %>%
  select(PLOT = Var1, AZC = Var2)%>%
  mutate(TRANSECT = paste0(PLOT, "-", AZC))

valid_transects = valid_transects$TRANSECT
#view(valid_transects)

head(fwd_diams)

#Adding a new column to fwd_diams for transect
fwd_diams = fwd_diams %>%
  mutate(PLOT = paste0(COMP, "-", 
                       stringr::str_pad(PLOT, width = 3, side = "left", pad = 0))) %>%
  mutate(TRANSECT = paste0(PLOT, "-", str_pad(AZ, width = 3, side = "left", pad = 0)), 
         AZ = as.numeric(AZ)) %>%
  select(COMP, PLOT, TRANSECT,  AZ, DIAM_mm)

#Checking for valid transects
fwd_diams %>%
  filter(!is.element(TRANSECT, valid_transects))

#Checking for missing transects
# A-023-000 transect ends at 18.1m, A-024-090 and C-022-270 transects ends at 17m
#believe data was being recorded at 19.5 m on transect; emailed crew to confirm;
# confirmed 7-27-2020: crew was recording fwd diams consistently at 19.5m after 
# the first day, where it was recorded at 2.5m and then either 9.5 or 19.5, probably
# depending on who did the sample. Might have to toss those first day locations.
valid_transects[!is.element(valid_transects,fwd_diams$TRANSECT)]



```
  
## Shrubs 
  - Distance_m, Maj_diam_m, minor_diam_m, height_m will be >=0 
  - species is the only column that can have an NA value, all other columns should have no NA values
  - COMPS wil be either A B or C 
  - status should be either L or D 
  - specific set of COMP/PLOT combinations that exist 
  - check for high outliers 
  - Location_m is >=0, and <=30 
  - distance_m for high outliers, >=0 
  - check species code is valid, look at unique values and fix any issues 
  - Major diam >= Minor Diam 

  
```{r}
head(shrubs)

#Distance_m, Maj_diam_m, minor_diam_m will be >=0 
#Maxium height of 25m seems wrong, as well max minor diam of 15.5m while max Major diam is 5.5
summary(shrubs)

shrubs %>%
  filter(DISTANCE_m<0| MAJ_DIAM_m<0 |MINOR_DIAM_m<0)

#check for high outliers in distance, check the row for 2.6m and make sure that it is a large shrub
shrubs %>%
  filter(DISTANCE_m>2)

#Identifying NA in species and making sure NA exists in no other columns, there are no NAs
shrubs %>%
  filter(is.na(SPECIES))

shrubs %>%
 filter(is.na(COMP) | is.na(PLOT) | is.na(LOCATION_m) | is.na(DISTANCE_m) | 
          is.na(MAJ_DIAM_m)| is.na(MINOR_DIAM_m)|is.na(HEIGHT_m) |  is.na(STATUS))

#COMP will be either A, B, or C 
valid_COMP <- 
  c("A", "B", "C")

shrubs %>%
  filter(!is.element(COMP, valid_COMP))

#Status will either be L or D 
valid_status <- 
  c("L", "D")

shrubs %>%
  filter(!is.element(STATUS, valid_status))

#specific COMP/PLOT/AZ
#view(valid_transects)
head(shrubs)
#Adding a new column  for transect
shrubs = shrubs %>%
  mutate(PLOT = paste0(COMP, "-", 
                       stringr::str_pad(PLOT, width = 3, side = "left", pad = 0))) %>%
  mutate(TRANSECT = paste0(PLOT, "-", str_pad(AZ, width = 3, side = "left", pad = 0)), 
         AZ = as.numeric(AZ)) %>%
  select(COMP, PLOT, TRANSECT, AZ, SPECIES, STATUS, LOCATION_m, DISTANCE_m, 
         MAJ_DIAM_m, MINOR_DIAM_m, HEIGHT_m)

#Checking for valid transects
shrubs %>%
  filter(!is.element(TRANSECT, valid_transects))

#Checking for missing transects 
#A- 001 - 000, A - 001 - 090, A- 001 - 180, A - 001 - 270, A- 005 0 270 ; A-021-090 
#All of these locations have no shrubs in the metadata; rechecked 11-3-20
valid_transects[!is.element(valid_transects,shrubs$TRANSECT)]

#Location_m is >=0, and <=30 
shrubs %>%
  filter(LOCATION_m<0 | LOCATION_m>30)

#check species code is valid, look at unique values and fix any issues 
unique(shrubs$SPECIES)

#Major diam >= Minor Diam 
#C- 010- 000 we are highly confident that the crew incorrectly entered the decimal place for the minor diameter so it reads as 15.5 but is actually 1.55; we plan to change this value 
#B-010-270, c-025-270, B-013-280: we are not confident why this test is not true so we will leave the data as is 
shrubs %>%
  filter(MAJ_DIAM_m<MINOR_DIAM_m)

#hand edit specific values
shrubs[shrubs$TRANSECT == "C-010-000" & shrubs$LOCATION_m == 5.15, "MINOR_DIAM_m"] = 1.55
shrubs[shrubs$TRANSECT == "C-010-000" & shrubs$LOCATION_m == 5.15, "MINOR_DIAM_m"]

```

## CWD

```{r}
# tests
## only valid combinations of COMP:PLOT:AZ
## location in 0:30
## DIAMs < DIAMi < DIAMl
## DIAMi >= 7.6
## DECAY in c(1, 2, 3, 4, 5)
## SPECIES in valid species
## high outliers in diam i
## high outliers in diam s
## high outliers in diam l
## high outliers in length

# only valid combinations of COMP:PLOT:AZ
#Adding a new column  for transect
cwd = cwd %>%
  mutate(PLOT = paste0(COMP, "-", 
                       stringr::str_pad(PLOT, width = 3, side = "left", pad = 0))) %>%
  mutate(TRANSECT = paste0(PLOT, "-", str_pad(AZ, width = 3, side = "left", pad = 0)), 
         AZ = as.numeric(AZ)) %>%
  select(COMP, PLOT, TRANSECT, AZ, SPECIES, DECAY, LOCATION_m, DIAM_I_cm, DIAM_S_cm, 
         DIAM_L_cm, LENGTH_m)

#Checking for valid transects
cwd %>%
  filter(!is.element(TRANSECT, valid_transects))

# Checking for missing transects 
# A-24-00: no CWD comment in plot metadata
# C-26-00: no CWD comment in plot metadata
# A-24-90: no CWD comment in plot metadata
valid_transects[!is.element(valid_transects,cwd$TRANSECT)]

# location in 0:30
cwd %>% filter(LOCATION_m < 0 | LOCATION_m > 30) # no fails

# DIAMs < DIAMi < DIAMl
cwd %>% filter(!(DIAM_S_cm <= DIAM_L_cm &
                 DIAM_S_cm <= DIAM_I_cm &
                 DIAM_I_cm <= DIAM_L_cm)) # 97 fails, break it up:

cwd %>% filter(!(DIAM_S_cm <= DIAM_I_cm)) # 26 obs where S > I
cwd %>% filter(!(DIAM_S_cm <= DIAM_L_cm)) # 3 obs where S > L
cwd %>% filter(!(DIAM_I_cm <= DIAM_L_cm)) # 73 obs where I > L

## i do remember some confusion during training about measuring diam_i always 
## orthogonal to the main axis, rather than along the transect. That could 
## account for diams at intersection > large end diams. 
## UPDATE: AFter conferring with crew, it sounds like it was fairly common for 
## pieces to be wider towards the center (where the cylinder is reasonably whole)
## than at the "large end" which was often damaged. These might complicate 
## carbon stock estimates using the large/small/length abbroach to volume estimation, 
## but shouldn't be a problem for estimating fuel loads from the intersection. 

ggplot(data = cwd,
       aes(x = DIAM_I_cm, y = DIAM_L_cm))+
  geom_point()+
  coord_fixed()+
  geom_abline(intercept = 0, slope = 1, color ='red') # IDK this looks mostly ok, 
## could just be regular measurement error? 

## the 3 obs where S > L look like just swappwed around columns, not a big 
## deal for carbon calculations. The idea that in 25% of the observations 
## diam at intersection > large end diameter is concerning, esp given the 
## confusion viz. orthogonal vs. along-transect diameters at intersection. 
## However, plotting the data makes it look like just normal measurement 
## error - the relationship between diam_i and diam_l is fuzzy in both directions.
## Not going to update any of these values now; check with Battles whether his 
## biomass equations need small end < large end or if they just need 
## diam for end 1 and end 2
## update; see note above its NBD

# DIAMi >= 7.6
cwd %>% filter(DIAM_I_cm < 7.6) # 1 fail, with diam == 2.8. Given the small 
# and large ends, i'm reasonably sure its a typo and supposed to be 20.8. 
cwd[cwd$COMP=='C'& cwd$PLOT=='10'&cwd$AZ=='0'&round(cwd$LOCATION_m,1)==20.7,'DIAM_I_cm'] = 20.8

# pretty sure the protocol calls for DIAMs and DIAMl to also be >= 7.6
cwd %>% filter(DIAM_S_cm < 7.6)
cwd %>% filter(DIAM_L_cm < 7.6)
## these all look like valid observations given the diameter at intersection; 
## i'm going to assume the crews messed up and meaured diamter of a broken 
## off bit instead of truncating at 7.6. leave as-is (because we don't know 
## how much to truncate the length)


# DECAY in c(1, 2, 3, 4, 5)
cwd %>% filter(!is.element(DECAY, c(1, 2, 3, 4, 5))) # no fails


# SPECIES in valid species
unique(cwd$SPECIES)

## make some more standardized species codes
valid_cwd_spp = c('ABCO', 'CADE', 'PILA', 'PIPO', 'PSME', 'NODE', 'QUKE',
                  'FIR', 'PINE', 'HARDWOOD', 'CONIFER', 'UNK')

## maap the standardized codes to teh existing ones
cwd_spp_mapping = 
  data.frame(map_from = c('UC', 'IC', 'FIR', 'CADE', 'U', 'PIPO', 'PSME', 
                          'UP', 'PILA', 'UF', 'ABCO', 'UH', 'NODE', 'QUKE'),
             map_to = c('CONIFER', 'CADE', 'FIR', 'CADE', 'UNK', 'PIPO', 'PSME',
                        'PINE', 'PILA', 'FIR', 'ABCO', 'HARDWOOD', 'NODE', 'QUKE')) %>%
  mutate(map_from = as.character(map_from),
         map_to = as.character(map_to))

## convert to the standardized codes
cwd = 
  cwd %>%
  left_join(.,
            cwd_spp_mapping,
            by = c('SPECIES' = 'map_from')) %>%
  mutate(SPECIES = map_to) %>%
  select(-map_to)
  


# high outliers in diam i
ggplot(data = cwd,
       aes(x = DIAM_I_cm))+
  geom_histogram()+
  geom_boxplot(aes(y = -7), width = 5)+
  theme_minimal() # looks fine; if any slightly under skewed

## high outliers in diam s
ggplot(data = cwd,
       aes(x = DIAM_S_cm))+
  geom_histogram()+
  geom_boxplot(aes(y = -7), width = 5)+
  theme_minimal() # more skewed but OK for CWD

## high outliers in diam l
ggplot(data = cwd,
       aes(x = DIAM_L_cm))+
  geom_histogram()+
  geom_boxplot(aes(y = -7), width = 5)+
  theme_minimal() # also all looks plausible, nothing over 1m

## high outliers in length
ggplot(data = cwd,
       aes(x = LENGTH_m))+
  geom_histogram()+
  geom_boxplot(aes(y = -7), width = 5)+
  theme_minimal() # looks good; 15m very long but plausible

```

## Trees

```{r}

# tests:
## only valid combinations of comp-plot
## location in valid domain (0 <= location <= 15)
## status is L or D
## DBH > 0
## if tree outside of the 1m sub-belt, DBH >= 11.43
## tree within a belt
## Height > 1.3
## if status is live, !isna(htcb) & (height > htcb) & (htcb > 0)
## if status is dead, !isna(decay) & decay in c(1, 2, 3, 4, 5)
## species in valid set
## TS in c(NW, NE, SW, SE, EN, ES, WN, WS)
## check for high outliers in DBH
## check for high outliers in height
## check for outliers (among live trees) of dbh:height relationship


head(trees)
# only valid combinations of comp-plot
#Adding a new column  for transect
trees = trees %>%
  mutate(PLOT = paste0(COMP, "-", 
                       stringr::str_pad(PLOT, width = 3, side = "left", pad = 0))) %>%
  mutate(TAG = as.character(TREE),
         TREEID = paste0(PLOT,
                         '-',
                         TS,
                         '-',
                         LOCATION,
                         '-',
                         DISTANCE)) %>%
  select(COMP, PLOT, TREEID, TAG, STATUS, SPECIES, TS, LOCATION, DISTANCE, 
         dbh_cm = DBH, height_m = HEIGHT, htcb_m = HTCB,
         decay = DECAY)

#Checking for valid plots
trees %>%
  filter(!is.element(PLOT, valid_plots))

# this tree is located in the middle of the trelist for B-26 in the source 
# datasheet; fix the typo:
trees[trees$COMP=='B'&trees$PLOT=='B-023'&trees$TREEID=='B-023-EN-9.7-0.7','PLOT'] = 'B-026'
trees[trees$COMP=='B'&trees$TREEID=='B-023-EN-9.7-0.7','TREEID'] = 'B-026-EN-9.7-0.7'
trees %>%
  filter(!is.element(PLOT, valid_plots))

#Checking for missing transects 
#A- 001 - 000, A - 001 - 090, A- 001 - 180, A - 001 - 270, A- 005 0 270 ; 
#All of these locations have no shrubs in the metadata up to date July 20, 2020
valid_plots[!is.element(valid_plots,trees$PLOT)]


# location in valid domain
trees %>% filter(LOCATION < 0 | LOCATION > 15) # no fails

# status is l or d
trees %>% filter(!is.element(STATUS, c('L', 'D'))) # no fails

# DBH > 0 
trees %>% filter(dbh_cm <= 0) # no fails

# tree within belt
trees %>%
  filter(
    # show trees that aren't in the belt, where in the belt means:
    !(
      # either (location <= 15 and distance <= 5), or 
      # (location <= 5 and distance <= 15)
      (LOCATION <= 15 & DISTANCE <= 5) | (LOCATION <= 5 & DISTANCE <= 15)
    )
  ) # no failures


# if tree not within sub-belt, DBH >= 11.4
trees %>% 
  
  filter(
    # show trees that aren't in the belt, where in the belt means:
    !(
      # either (location <= 15 and distance <= 5), or 
      # (location <= 5 and distance <= 15)
      (LOCATION <= 15 & DISTANCE <= 1) | (LOCATION <= 1 & DISTANCE <= 15)
    ) & dbh_cm < 11.4
    
  ) 
# 1 fail; either distance is a typo 
## (should be 0.19?), DBH is a typo (should be 19?) or the crew accidentally 
## included a tree which was too far away. Without other info, take the 
## data at face value (tree is too small & too far off transect to be included) 
## so drop it:
trees = 
  trees %>% 
  filter(# keep only trees which are in the sub-belt, or geq 11.4cm
        ((LOCATION <= 15 & DISTANCE <= 1) | (LOCATION <= 1 & DISTANCE <= 15)) | 
          dbh_cm >= 11.4)


# height > 1.4
trees %>% filter(height_m < 1.4) # no fails

# if status is live, !isna(htcb) & (height > htcb) & (htcb > 0)
trees %>% filter(STATUS == 'L' & ((is.na(htcb_m))| !(height_m > htcb_m) | !(htcb_m > 0)))

# four trees with decay and no HTCB, but marked as live. These are almost definitely 
# dead trees with typo status. fix the statuses:
trees[trees$STATUS=='L' & is.na(trees$htcb_m) & !is.na(trees$decay), 'STATUS'] = 'D'

# if status is dead, !isna(decay) & decay in c(1, 2, 3, 4, 5)
trees %>% filter(STATUS == 'D' & (!is.element(decay, c(1, 2, 3, 4, 5)))) # no fails

# species in valid set
unique(trees$SPECIES)
valid_tree_spp = c('ABCO', 'CADE', 'QUKE', 'PILA', 'PIPO', 
                   'PSME', 'NODE', 'SEGI', 
                   'ARME', 'CHCH', 'CONU')

trees %>% filter(!is.element(SPECIES, valid_tree_spp)) # one 'A' and a bunch of 'U'
# 'U' is clearly unknown (and all of those are decayed snags). Not sure what 
# 'A' is; probably ABCO but could be ARME? No comment in metadata. Row above 
# it in the source spreadsheet is an ABCO, so either the crew expected it to 
# autofill (and it didnt) or it tried to autofill and they deleted it to 
# code ARME...
# going to assume ABCO
trees[trees$COMP=='C' & trees$PLOT=='22' & trees$LOCATION == 9.6 & trees$DISTANCE==0.2,
      'SPECIES'] = 'ABCO'

# TS in c(NW, NE, SW, SE, EN, ES, WN, WS)
trees %>% filter(!is.element(TS, c('NW', 'NE', 'SW', 'SE', 'EN', 'ES', 'WN', 'WS')))


# check for high outliers in DBH
ggplot(data = trees, 
       aes(x = dbh_cm))+
  geom_histogram()+
  geom_boxplot(aes(y = -10),
               width = 7)+
  theme_minimal() # looks fine

# check for high outliers in height
ggplot(data = trees, 
       aes(x = height_m))+
  geom_histogram()+
  geom_boxplot(aes(y = -10),
               width = 7)+
  theme_minimal() # looks fine


## check for outliers (among live trees) of dbh:height relationship
ggplot(data = trees %>% filter(STATUS=='L'), 
       aes(x = dbh_cm, y = height_m, color = SPECIES))+
  geom_point()+
  theme_minimal()+
  geom_abline(intercept = 0, slope = 0.2, lty = 'dashed', color = 'red')+
  scale_color_viridis_d() # a few samples with very low height for their DBH

trees %>% 
  filter(STATUS == 'L' & height_m < 5 & dbh_cm > 20)
# low heights for hardwoods make sense, leave as-is


```


## Presence/absence

```{r}
head(pres_abs)
head(pres_abs_2)

# first off, join the tables
pres_abs = 
  rbind(pres_abs,
        pres_abs_2)

# tests
## valid combinations of comp/plot/az
## valid combinations of comp/plot/az/location
## location in valid range
## LD/FWD/CWD/SHRUB either 0 or 1 (no NAs)
## CANOPY either NA, 0, or 1

# fix a lowercase c comp which was discovered below:
pres_abs = pres_abs %>% mutate(COMP = ifelse(COMP=='c', 'C', COMP))

## ditch the all-NA rows; crews copypasted location template too many times
## don't ditch the observations from b-8-270 where they didn't copy the ID
## info...
pres_abs = pres_abs %>% filter(!(is.na(LD)&is.na(FWD)&is.na(CWD)&is.na(SHRUB)))

## for the 290 remaining rows, these are all from B-8-270
pres_abs = 
  pres_abs %>%
  mutate(COMP = ifelse(is.na(COMP)&is.na(PLOT)&is.na(AZ),
                       'B',
                       COMP),
         PLOT = ifelse(is.na(PLOT)&is.na(AZ),
                       '8',
                       PLOT),
         AZ = ifelse(is.na(AZ),
                       '270',
                       AZ))

pres_abs = pres_abs %>%
  mutate(PLOT = paste0(COMP, "-", 
                       stringr::str_pad(PLOT, width = 3, side = "left", pad = 0))) %>%
  mutate(TRANSECT = paste0(PLOT, "-", str_pad(AZ, width = 3, side = "left", pad = 0)), 
         AZ = as.numeric(AZ), 
         Displacement_m = LOCATION, 
         LOCATION = paste0(TRANSECT, "-", LOCATION)) %>%
  select(COMP, PLOT, TRANSECT, LOCATION, AZ, Displacement_m, 
         LD, FWD, CWD, SHRUB, CANOPY)

#confirm all of our LOCATION - All of the correct combinations of COMP, PLOT, TRANSECT, LOCATION
valid_plots = c("A-001", "A-003", "A-005", "A-023", "A-025", "A-020", "A-024", 
                "A-028", "B-008", "B-013", "B-010", "B-016", "B-026", "C-010", 
                "C-011", "C-025", "C-026", "C-022")

valid_locations <- 
  expand.grid(valid_plots, 
              c("000", "090", "180", "270"), 
              c(seq(from = 0, to = 29.9, by = 0.1))) %>%
  select(PLOT = Var1, TRANSECT = Var2, Displacement_m = Var3) %>%
  mutate(LOCATION = paste0(PLOT, "-", TRANSECT, "-", Displacement_m))

#Selecting the location column from the table to use 
#view(valid_locations)
valid_locations = valid_locations$LOCATION

#Checking for extra locations 
pres_abs %>%
  filter(!is.element(LOCATION, valid_locations))

#Checking for missing locations 
valid_locations[!is.element(valid_locations,pres_abs$LOCATION)]
# theres are all out-of-bounds locations


# checking allowable values; should be 0, 1, or NA. There will be some 
# NAs in the various surface columns on locations which were obstructed. 
# Lots of NAs expected inthe canopy column because only measured 1 per meter 
# (1 per 10 rows)
summary(pres_abs %>% select(LD, FWD, CWD, SHRUB, CANOPY))

# there's '11's in LD and FWD which should definitely be 1s;
# also call everything > 0 in CANOPY a 1
# and convert all columns to logical
pres_abs = 
  pres_abs %>% 
  mutate(LD = ifelse(LD>0, TRUE, FALSE),
         FWD = ifelse(FWD>0, TRUE, FALSE),
         CWD = ifelse(CWD>0, TRUE, FALSE),
         SHRUB = ifelse(SHRUB>0, TRUE, FALSE),
         CANOPY = ifelse(CANOPY>0, TRUE, FALSE))
summary(pres_abs %>% select(LD, FWD, CWD, SHRUB, CANOPY))

```

# Write outputs

```{r}
write.csv(litter_duff,
          here::here('02-data', '01-cleaned', 'litter_duff.csv'),
          row.names = FALSE)

write.csv(fwd_tallies,
          here::here('02-data', '01-cleaned', 'fwd_tallies.csv'),
          row.names = FALSE)


write.csv(fwd_diams,
          here::here('02-data', '01-cleaned', 'fwd_diams.csv'),
          row.names = FALSE)


write.csv(shrubs,
          here::here('02-data', '01-cleaned', 'shrubs.csv'),
          row.names = FALSE)


write.csv(cwd,
          here::here('02-data', '01-cleaned', 'cwd.csv'),
          row.names = FALSE)


write.csv(trees,
          here::here('02-data', '01-cleaned', 'trees.csv'),
          row.names = FALSE)


write.csv(pres_abs,
          here::here('02-data', '01-cleaned', 'pres_abs.csv'),
          row.names = FALSE)


write.csv(plot_metadata,
          here::here('02-data', '01-cleaned', 'plot_metadata.csv'),
          row.names = FALSE)

```


#Data Exploration 
```{r eval = FALSE}
#Zurr graphs for littler/duff, FWD tallies, FWD diam, CWD or just focus on 1 or 2 tables and do good job 

#Litter and Duff
head(litter_duff)
    
ggplot(litter_duff, aes(x=LOCATION, y =DUFF_cm)) +
  geom_point(aes(col=COMP)) + 
  labs(y="Duff depth (cm)", x= "Location")

ggplot(litter_duff, aes(x=as.factor(LOCATION), y =DUFF_cm)) +
  geom_boxplot() + 
  labs(y="Duff depth (cm)", x= "Location")

ggplot(litter_duff, aes(x =DUFF_cm)) +
  geom_histogram()

ggplot(litter_duff, aes(x=LOCATION, y =LITTER_cm)) +
  geom_boxplot(aes(col=COMP)) + 
  labs(y="Litter depth (cm)", x= "Location")

ggplot(litter_duff, aes(x=LOCATION, y =LITTER_cm)) +
  geom_point(aes(col=COMP)) + 
  labs(y="Litter depth (cm)", x= "Location")

ggplot(litter_duff, aes(x =LITTER_cm)) +
  geom_histogram()

#FWD Tallies
head(fwd_tallies_long)
head(fwd_tallies)

ggplot(fwd_tallies, aes(x=LOCATION, y = a1h)) + 
  geom_boxplot(aes(col=COMP)) + 
  labs(y="A1h", x = "Location")

ggplot(fwd_tallies, aes(x=LOCATION, y = a1h)) + 
  geom_point(aes(col=COMP))

ggplot(fwd_tallies, aes(x=a1h)) +
  geom_histogram()

ggplot(fwd_tallies_long, aes(x=LOCATION, y = count)) + 
  geom_boxplot(aes(col=fuel_type)) + 
  labs(y="Count", x = "Location")

ggplot(fwd_tallies_long, aes(x=LOCATION, y = count)) + 
  geom_point(aes(col=fuel_type))

ggplot(fwd_tallies_long, aes(x =count)) +
  geom_histogram()


```


##Scratch work 

```{r eval = FALSE}
#base R way to delete columns, everything before common is which rows to select and everything after common is which columns to select 

#this is how to pick which rows
fwd_tallies.A = 
  fwd_tallies[fwd_tallies$COMP == "A",]

#this is how to pick which columns
#fwd_tallies.clean = 
  #fwd_tallies[, c("COMP", "PLOT", "AZ", "LOCATION", #"a1h", "a10h", "a100h", "b1h", "b10h", "b100h")]

#tidyverse way to pick columns/rows 
fwd_tallies.B = 
  filter(fwd_tallies, COMP =="B")

fwd_tallies.B2 = 
  select(fwd_tallies.B, c(a1h, a10h, a100h))
head(fwd_tallies.B2)

fwd_tallies.B3 = 
  fwd_tallies %>% 
  filter(COMP == "B") %>%
  select(c(a1h, a10h, a100h))

head(fwd_tallies.B3)
#pipes try to automatically figure out which argument you are calling to, you can direct with a period 

head(cwd)

fwd_tallies_long <- 
  pivot_longer(fwd_tallies.clean, cols= c("a1h", "a10h", "a100h", "b1h", "b10h", "b100h"), names_to = "fuel_type", values_to = "count")

head(fwd_tallies_long)
#head(cwd_location,100)
#cwd_location <- 
  #cwd %>% 
  #select(COMP, PLOT, AZ) %>%
  #expand(nesting(COMP, PLOT), AZ)
  
#cwd %>%
#group_by(COMP, PLOT, AZ) %>%
#arrange(LOCATION_m, .by_group = TRUE) %>%
#summarise(ssd_cm2 = sum((DIAM_I_cm)^2)) %>% 
#ungroup()
```

